---
title: "Rossmann Store Sales - Demand Forecasting"
author name: "Obaid"
date: "2026-05-08"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
    highlight: tango
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Executive Summary
This project builds a demand forecasting model using the 
Rossmann Store Sales dataset. Four models were developed and 
compared progressively, from a naive store-average baseline 
through linear regression, decision trees, and random forests. 
The best-performing model, linear regression with store-level 
fixed effects, achieved an RMSE of approximately 1,370, 
reducing prediction error by 28% over the baseline.

\newpage

## 1 - Introduction

*"Retail is Detail!"*

Retail demand forecasting is one of the most practically valuable 
applications of machine learning in supply chain management. 
Knowing how much stock a store will need, days, weeks, or months 
in advance, directly impacts inventory levels, staffing decisions, 
and supplier orders.

This project uses the Rossmann Store Sales dataset, sourced 
from a Kaggle competition hosted by Rossmann, one of Europe's 
largest drug store chains. The dataset contains 1,017,209 
daily sales records across 1,115 stores spanning from 
January 2013 to July 2015. Each record includes information about 
promotions, holidays, and store-level characteristics such as 
store type, assortment level, and competitor proximity.

The objective is to predict daily sales (`Sales`) for each 
store-date combination. Model performance is evaluated using 
Root Mean Square Error (RMSE):

$$RMSE = \sqrt{\frac{1}{N}\sum_{i=1}^{N}(\hat{y}_i - y_i)^2}$$

A lower RMSE indicates better predictive accuracy.

The report is structured as follows: Section 2 describes the 
analytical approach, Section 3 presents exploratory data analysis, 
Section 4 covers feature engineering, Section 5 details the 
modelling process, Section 6 presents results, and Section 7 
concludes with limitations and future directions.

```{r load-libraries-data}
if(!require(tidyverse))    install.packages("tidyverse")
if(!require(caret))        install.packages("caret")
if(!require(lubridate))    install.packages("lubridate")
if(!require(scales))       install.packages("scales")
if(!require(knitr))        install.packages("knitr")
if(!require(rpart))        install.packages("rpart")
if(!require(rpart.plot))   install.packages("rpart.plot")
if(!require(randomForest)) install.packages("randomForest")


library(tidyverse)
library(caret)
library(lubridate)
library(scales)
library(knitr)
library(rpart)
library(rpart.plot)
library(randomForest)

# Unzip data if not already done
if(!file.exists("Data/train.csv")) {
  unzip("Data.zip", exdir = "Data")
}

train <- read.csv("Data/train.csv", stringsAsFactors = FALSE)
store <- read.csv("Data/store.csv", stringsAsFactors = FALSE)

rossmann <- train %>%
  left_join(store, by = "Store")

rossmann <- rossmann %>%
  mutate(Date = as.Date(Date))
```

```{r sanity-check}
data.frame(
  Metric = c("Total Rows", "Total Columns", "Date Range"),
  Value  = c(
    format(nrow(rossmann), big.mark = ","),
    ncol(rossmann),
    paste(min(rossmann$Date), "to", max(rossmann$Date))
  )
) %>% kable(caption = "Dataset Overview")
```

## 2 - Data Familiarisation

Before any analysis, it is important to understand what each column represents and what condition the data is in.

### 2.1 - Column Descriptions

The merged dataset contains 18 columns drawn from two source files.


| Column | Type | Description |
|---|---|---|
| Store | Integer | Unique store identifier (1–1115) |
| DayOfWeek | Integer | Day of week (1=Monday, 7=Sunday) |
| Date | Date | Calendar date of the record |
| Sales | Integer | Total sales for that day |
| Customers | Integer | Number of customers that day |
| Open | Binary | 1 = store open, 0 = store closed |
| Promo | Binary | 1 = promotion running that day |
| StateHoliday | Character | a=public, b=Easter, c=Christmas, 0=none |
| SchoolHoliday | Binary | 1 = schools closed that day |
| StoreType | Character | Store format (a, b, c, d) |
| Assortment | Character | Product range: a=basic, b=extra, c=extended |
| CompetitionDistance | Integer | Metres to nearest competitor store |
| CompetitionOpenSinceMonth | Integer | Month competitor opened |
| CompetitionOpenSinceYear | Integer | Year competitor opened |
| Promo2 | Binary | 1 = store runs a recurring promotion |
| Promo2SinceWeek | Integer | Week store joined Promo2 |
| Promo2SinceYear | Integer | Year store joined Promo2 |
| PromoInterval | Character | Months Promo2 is active e.g. "Jan,Apr,Jul,Oct" |

### 2.2 - First Look at the Data

```{r data-glimpse}
rossmann %>%
  head(6) %>%
  select(Store, DayOfWeek, Date, Sales, Customers,
         Open, Promo, StoreType, Assortment,
         CompetitionDistance) %>%
  kable(caption = "First Six Rows of the Dataset (Key Columns)")
```


### 2.3 - Missing Values

```{r missing-values}
missing_df <- data.frame(
  Column         = names(colSums(is.na(rossmann))),
  Missing_Values = as.integer(colSums(is.na(rossmann)))
) %>% 
  filter(Missing_Values > 0) %>%
  arrange(desc(Missing_Values))

kable(missing_df, caption = "Columns With Missing Values")
```

The missing values above are all in store metadata columns, not 
in the daily sales records themselves. This means every sales 
observation is intact. The missing metadata relates to competition 
and promotion history, which we will handle carefully in the 
feature engineering section.

## 3 - Exploratory Data Analysis

Before modelling, the training data was examined to uncover 
patterns in sales across time, promotions, store types, and 
competitor proximity. Closed store days are excluded from all 
analysis, a store recording zero sales because it is shut is 
not a demand signal.

### 3.1 - Sales Distribution

```{r plot-sales-dist}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  ggplot(aes(x = Sales)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Distribution of Daily Sales (Open Stores Only)") +
  xlab("Daily Sales (Euros)") + ylab("Count")
```

Daily sales follow a right-skewed distribution, with most 
store-days generating between 3,000 and 10,000 euros. A small 
number of exceptionally high-performing days pull the tail 
rightward. This skew is typical in retail sales data and suggests 
that a small number of stores or calendar events drive 
disproportionate revenue, a pattern familiar in demand planning.

### 3.2 - Effect of Promotions

```{r plot-promo}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  group_by(Promo) %>%
  summarize(
    Avg_Sales    = round(mean(Sales), 0),
    Median_Sales = round(median(Sales), 0),
    Count        = n()
  ) %>%
  mutate(Promo = ifelse(Promo == 1, "Promotion Active",
                                    "No Promotion")) %>%
  kable(caption = "Sales With vs Without Promotion")

rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(Promo = ifelse(Promo == 1, "Promotion Active",
                                    "No Promotion")) %>%
  ggplot(aes(x = Promo, y = Sales, fill = Promo)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Sales: Promotion vs No Promotion") +
  xlab("") + ylab("Daily Sales (Euros)") +
  theme(legend.position = "none")
```

Promotions have a clear and consistent positive effect on sales. 
This is the strongest single signal in the dataset. From a supply 
chain perspective, this reinforces the importance of promotion 
calendars in demand planning, as promotional uplifts must be 
anticipated well in advance to avoid stockouts during campaign 
periods.

### 3.3 - Day of Week Patterns

```{r plot-dow}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  group_by(DayOfWeek) %>%
  summarize(avg_sales = mean(Sales)) %>%
  ggplot(aes(x = DayOfWeek, y = avg_sales)) +
  geom_col(fill = "darkorange", color = "black") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = 1:7,
    labels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")) +
  ggtitle("Average Sales by Day of Week") +
  xlab("Day of Week") + ylab("Average Sales (Euros)")
```

Sales peak on Mondays and Fridays and drop on Sundays, 
reflecting German retail trading laws that restrict Sunday 
opening hours. This strong day-of-week pattern is a fundamental 
feature in retail demand forecasting. Any model that ignores 
which day of the week it is will perform poorly.

### 3.4 - Monthly Seasonality

```{r plot-month}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  group_by(Month) %>%
  summarize(avg_sales = mean(Sales)) %>%
  ggplot(aes(x = Month, y = avg_sales)) +
  geom_col(fill = "seagreen", color = "black") +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Average Sales by Month") +
  xlab("Month") + ylab("Average Sales (Euros)")
```

December shows the highest average sales, consistent with 
Christmas shopping behaviour. A secondary peak is visible in 
spring. These seasonal patterns are well understood in retail 
demand planning, they confirm that month of year is a necessary 
feature in any forecasting model.

### 3.5 - Sales Trend Over Time

```{r plot-trend}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  group_by(Date) %>%
  summarize(avg_sales = mean(Sales)) %>%
  ggplot(aes(x = Date, y = avg_sales)) +
  geom_line(color = "darkblue", alpha = 0.6) +
  geom_smooth(method = "loess", color = "red", span = 0.1) +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Average Daily Sales Over Time") +
  xlab("Date") + ylab("Average Sales (Euros)")
```

The overall sales trend is slightly upward across the two and a 
half year period, with strong recurring seasonal spikes each 
December. Short-term volatility reflects weekly cycles and 
promotional activity. The smooth trend line confirms steady 
underlying growth across the observation window.

### 3.6 - Store Type Comparison

```{r plot-storetype}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  group_by(StoreType) %>%
  summarize(Avg_Sales = round(mean(Sales), 0), Count = n()) %>%
  kable(caption = "Average Sales by Store Type")

rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  ggplot(aes(x = StoreType, y = Sales, fill = StoreType)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Sales Distribution by Store Type") +
  xlab("Store Type") + ylab("Daily Sales (Euros)") +
  theme(legend.position = "none")
```

Store type B stands out with significantly higher average sales 
than types A, C, and D. This suggests store format is a strong 
predictor of baseline sales volume, likely reflecting differences 
in store size, location type, and product range. Store type will 
be included as a categorical feature in all models.

### 3.7 - Competitor Distance Effect

```{r plot-competitor}
rossmann %>%
  filter(Open == 1, Sales > 0, !is.na(CompetitionDistance)) %>%
  mutate(dist_bucket = cut(CompetitionDistance,
    breaks = c(0, 500, 1000, 2000, 5000, 10000, Inf),
    labels = c("<500m","500m-1km","1-2km",
               "2-5km","5-10km",">10km"))) %>%
  group_by(dist_bucket) %>%
  summarize(avg_sales = mean(Sales)) %>%
  ggplot(aes(x = dist_bucket, y = avg_sales)) +
  geom_col(fill = "purple", color = "black") +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Average Sales by Competitor Distance") +
  xlab("Distance to Nearest Competitor") +
  ylab("Average Sales (Euros)")
```

Stores with closer competitors tend to generate higher sales 
rather than lower. This counterintuitive finding likely reflects 
a location effect: stores in dense urban areas attract more 
competitors but also more customers. Competitor proximity is 
therefore a proxy for footfall rather than a direct negative 
pressure on sales.

### 3.8 - Competitor Distance: Controlling for Customers

The previous chart showed higher sales near competitors, but 
this could simply reflect busier locations. To separate the 
location effect from the competition effect, we control for 
customer numbers by examining sales per customer and 
average customer count across competitor distance bands.

If avg_customers drops near competitors but sales_per_customer 
stays flat, the location effect explains everything. If 
sales_per_customer also drops near competitors, there is a 
genuine competitive pricing or range pressure effect.

```{r plot-competitor-controlled}
rossmann %>%
  filter(Open == 1, Sales > 0, Customers > 0,
         !is.na(CompetitionDistance)) %>%
  mutate(
    sales_per_customer = Sales / Customers,
    dist_bucket = cut(CompetitionDistance,
      breaks = c(0, 500, 1000, 2000, 5000, 10000, Inf),
      labels = c("<500m","500m-1km","1-2km",
                 "2-5km","5-10km",">10km"))) %>%
  group_by(dist_bucket) %>%
  summarize(
    avg_sales_per_customer = round(mean(sales_per_customer), 2),
    avg_customers          = round(mean(Customers), 0),
    avg_sales              = round(mean(Sales), 0)
  ) %>%
  kable(caption = "Sales Per Customer by Competitor Distance")
```


The results reveal two competing effects. Stores near competitors 
attract significantly more customers, nearly 50% more footfall 
compared to stores with no nearby competitor. This confirms the 
location effect hypothesis: competitor proximity is a marker of 
dense, high-traffic areas.

However, sales per customer tell a different story. Customers 
spend approximately 8 euros per visit at stores with competitors 
within 500 metres, compared to over 10 euros at stores with no 
nearby competitor. This suggests a genuine basket splitting 
effect, customers in competitive areas distribute their 
purchases across multiple stores rather than consolidating at one.

From a supply chain perspective, these are fundamentally 
different store profiles despite similar total sales figures. 
High-footfall, low-basket stores require higher replenishment 
frequency and smaller order quantities, while isolated stores 
with larger baskets benefit from less frequent but larger 
deliveries. A single forecasting model must capture both 
behaviours.

### 3.9 - State Holiday Effect

```{r plot-stateholiday}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(StateHoliday = case_when(
    StateHoliday == "a" ~ "Public Holiday",
    StateHoliday == "b" ~ "Easter Holiday",
    StateHoliday == "c" ~ "Christmas",
    TRUE                ~ "No Holiday"
  )) %>%
  group_by(StateHoliday) %>%
  summarize(
    Avg_Sales    = round(mean(Sales), 0),
    Median_Sales = round(median(Sales), 0),
    Count        = n()
  ) %>%
  kable(caption = "Sales by State Holiday Type")

rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(StateHoliday = case_when(
    StateHoliday == "a" ~ "Public Holiday",
    StateHoliday == "b" ~ "Easter Holiday",
    StateHoliday == "c" ~ "Christmas",
    TRUE                ~ "No Holiday"
  )) %>%
  ggplot(aes(x = StateHoliday, y = Sales, fill = StateHoliday)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Sales Distribution by State Holiday Type") +
  xlab("") + ylab("Daily Sales (Euros)") +
  theme(legend.position = "none")
```

State holidays show a notable effect on sales patterns. Most 
stores close on public holidays, those that remain open show 
distinctly different sales behaviour depending on the holiday 
type. Christmas and Easter periods tend to drive elevated sales 
in the days surrounding the holiday, making holiday calendar 
awareness essential in retail demand planning.

### 3.10 - School Holiday Effect

```{r plot-schoolholiday}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  group_by(SchoolHoliday) %>%
  summarize(
    Avg_Sales    = round(mean(Sales), 0),
    Median_Sales = round(median(Sales), 0),
    Count        = n()
  ) %>%
  mutate(SchoolHoliday = ifelse(SchoolHoliday == 1,
                                "School Holiday",
                                "Normal Day")) %>%
  kable(caption = "Sales: School Holiday vs Normal Day")

rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(SchoolHoliday = ifelse(SchoolHoliday == 1,
                                "School Holiday",
                                "Normal Day")) %>%
  ggplot(aes(x = SchoolHoliday, y = Sales, fill = SchoolHoliday)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Sales: School Holiday vs Normal Day") +
  xlab("") + ylab("Daily Sales (Euros)") +
  theme(legend.position = "none")
```

School holidays show a modest effect on daily sales. This likely 
reflects changed shopping patterns when families have more 
flexibility during the day, particularly relevant for drug stores 
which serve a broad customer demographic including parents and 
children. Including school holiday status as a model feature 
captures this behavioural shift.

### 3.11 - Assortment Effect

```{r plot-assortment}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(Assortment = case_when(
    Assortment == "a" ~ "Basic",
    Assortment == "b" ~ "Extra",
    Assortment == "c" ~ "Extended"
  )) %>%
  group_by(Assortment) %>%
  summarize(
    Avg_Sales    = round(mean(Sales), 0),
    Median_Sales = round(median(Sales), 0),
    Count        = n()
  ) %>%
  kable(caption = "Sales by Assortment Level")

rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(Assortment = case_when(
    Assortment == "a" ~ "Basic",
    Assortment == "b" ~ "Extra",
    Assortment == "c" ~ "Extended"
  )) %>%
  ggplot(aes(x = Assortment, y = Sales, fill = Assortment)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Sales Distribution by Assortment Level") +
  xlab("Assortment Level") + ylab("Daily Sales (Euros)") +
  theme(legend.position = "none")
```

Assortment level shows a clear relationship with sales volume. 
Stores with an extended product range tend to generate higher 
sales, which makes intuitive sense, more products available 
means larger potential basket sizes. This also has direct 
implications for inventory management: extended assortment stores 
require broader SKU coverage and more complex replenishment 
planning than basic assortment stores.

### 3.12 - Promo2 Effect

```{r plot-promo2}
rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  group_by(Promo, Promo2) %>%
  summarize(
    Avg_Sales = round(mean(Sales), 0),
    Count     = n(),
    .groups   = "drop"
  ) %>%
  mutate(
    Promo  = ifelse(Promo  == 1, "Promo Active",  "No Promo"),
    Promo2 = ifelse(Promo2 == 1, "Promo2 Active", "No Promo2")
  ) %>%
  kable(caption = "Sales by Promo and Promo2 Combination")

rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(
    Promo  = ifelse(Promo  == 1, "Promo Active",  "No Promo"),
    Promo2 = ifelse(Promo2 == 1, "Promo2 Active", "No Promo2")
  ) %>%
  ggplot(aes(x = Promo, y = Sales, fill = Promo2)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Sales by Promo and Promo2 Combination") +
  xlab("") + ylab("Daily Sales (Euros)") +
  labs(fill = "Promo2 Status")
```

Surprisingly, stores participating in Promo2 show slightly lower 
median sales than non-participating stores, both with and without 
a standard promotion active. This counterintuitive finding likely 
reflects a store selection effect: Rossmann may enrol 
lower-performing stores into the recurring promotion programme 
to stimulate growth, meaning Promo2 stores were already weaker 
before the promotion began. Alternatively, it may indicate 
promotion fatigue, customers becoming desensitised to 
recurring discounts over time, a well-documented pattern in 
retail demand planning.


Feature selection was informed by both statistical analysis and 
domain knowledge. School holidays, state holidays, and the Promo2 
flag were excluded despite showing some signal in the EDA. 
SchoolHoliday showed a modest and inconsistent effect insufficient 
to justify the added complexity. StateHoliday affected too few 
observations to produce reliable model estimates. Promo2 was 
excluded on theoretical grounds, the recurring promotion is 
primarily a slow inventory clearance mechanism rather than a 
genuine demand driver, meaning its negative correlation with 
sales reflects store selection bias rather than a causal 
relationship.

## 4 - Feature Engineering

Before modelling, the raw dataset was transformed into a clean 
feature matrix. Three steps were applied.

### 4.1 - Removing Closed Store Days

Observations where stores were closed (Sales = 0) were removed. 
A store recording zero sales because it is shut is not a demand 
signal, including these would bias the model toward 
underpredicting on open days.

### 4.2 - Date Feature Extraction

The raw `Date` column was decomposed into two numeric features:

- **Month** — captures seasonal patterns (December peak, 
spring uplift)
- **Year** — captures the slight upward sales trend over time

The original Date column is then discarded as Month and Year 
together carry all the temporal information our models need.

### 4.3 - Handling Missing Values

`CompetitionDistance` contained missing values for a subset of 
stores. These were imputed with the median competition 
distance across all stores. Stores with no recorded competitor 
are likely in remote or rural locations, the median provides a 
conservative, robust estimate without introducing extreme values.

### 4.4 - Categorical Encoding

The following variables were converted to factors so that 
tree-based models can treat them correctly as categorical 
variables rather than continuous numbers:

- Store, StoreType, Assortment
- DayOfWeek, Promo, Month, Year

### 4.5 - Final Feature Matrix

```{r feature-engineering}
rossmann_model <- rossmann %>%
  filter(Open == 1, Sales > 0) %>%
  mutate(
    Month = month(Date),
    Year  = year(Date)
  ) %>%
  mutate(CompetitionDistance = ifelse(
    is.na(CompetitionDistance),
    median(CompetitionDistance, na.rm = TRUE),
    CompetitionDistance
  )) %>%
  mutate(
    Store      = as.factor(Store),
    StoreType  = as.factor(StoreType),
    Assortment = as.factor(Assortment),
    DayOfWeek  = as.factor(DayOfWeek),
    Promo      = as.factor(Promo),
    Month      = as.factor(Month),
    Year       = as.factor(Year)
  ) %>%
  select(Sales, Store, DayOfWeek, Promo,
         StoreType, Assortment,
         CompetitionDistance, Month, Year, Date)

glimpse(rossmann_model)
cat("Rows after filtering:", nrow(rossmann_model), "\n")
cat("Any missing values:", anyNA(rossmann_model), "\n")
```

The final feature matrix contains 844,338 observations and 
9 columns, 1 target variable and 8 features. Approximately 17% 
of the original records were removed as closed store days, 
leaving only genuine trading days for model training. No missing 
values remain.

## 5 - Modelling

### 5.1 - Train/Test Split

In demand forecasting, models must always be validated on future 
data, training on the past and testing on the future mirrors 
real operational conditions. A random split would allow the model 
to train on December 2014 data and test on January 2014, which 
would be unrealistic and overoptimistic.

The dataset was therefore split by time:

- **Training set** — all observations before 19 June 2015
- **Test set** — the final 6 weeks of data (19 June to 31 July 2015)

This ensures the model is evaluated on genuinely unseen future 
data, providing a realistic estimate of forecasting performance.

```{r train-test-split}
split_date <- as.Date("2015-06-19")

train_set <- rossmann_model %>%
  filter(Date < split_date) %>%
  select(-Date)

test_set <- rossmann_model %>%
  filter(Date >= split_date) %>%
  select(-Date)

cat("Training rows:", nrow(train_set), "\n")
cat("Test rows:"    , nrow(test_set),  "\n")

RMSE <- function(true, predicted){
  sqrt(mean((true - predicted)^2))
}
```

The training set contains 802,942 observations covering 
January 2013 to June 2015. The test set contains 41,396 
observations representing the final 6 weeks of trading data 
from June to July 2015.

### 5.2 - Baseline Model

Before applying machine learning, a simple baseline was 
established by predicting each store's average historical 
sales from the training set. This requires no feature 
information, only the store identifier.

Any model that cannot outperform this baseline is not adding 
value beyond what a simple lookup table could provide.

```{r model-baseline}
store_avgs <- train_set %>%
  group_by(Store) %>%
  summarize(pred = mean(Sales))

baseline_pred <- test_set %>%
  left_join(store_avgs, by = "Store") %>%
  pull(pred)

rmse_baseline <- RMSE(test_set$Sales, baseline_pred)
cat("Baseline RMSE:", rmse_baseline, "\n")

rmse_results <- tibble(
  Model = "Baseline (Store Average)",
  RMSE  = rmse_baseline
)
kable(rmse_results, caption = "Model Performance Tracking")
```

The baseline RMSE of 1,916 represents the floor all 
subsequent models must beat. Being off by approximately 1,900 
euros on a typical daily sales figure of 5,000-7,000 euros 
represents an error of roughly 30%, the direct consequence 
of ignoring every demand driver identified in the EDA. This 
gives us confidence there is substantial room for improvement.

### 5.3 - Linear Regression

Linear regression was applied as the first proper machine 
learning model. It assumes each feature contributes 
independently and additively to sales, learning a separate 
coefficient for each variable in the feature matrix.

$$\hat{Sales} = \beta_0 + \beta_1 \cdot Store + \beta_2 \cdot DayOfWeek + ... + \epsilon$$

With 1,115 unique stores encoded as factors, the model learns 
a baseline sales level for each store and adjusts upward or 
downward based on promotions, seasonality, assortment, and 
competitor distance.

Predicted values were clipped at zero, daily sales cannot 
be negative, and linear models can occasionally produce 
negative predictions for low-sales scenarios.

```{r model-lm}
lm_model <- lm(Sales ~ ., data = train_set)

lm_pred <- predict(lm_model, newdata = test_set)
lm_pred <- pmax(lm_pred, 0)

rmse_lm <- RMSE(test_set$Sales, lm_pred)
cat("Linear Regression RMSE:", rmse_lm, "\n")

rmse_results <- bind_rows(rmse_results,
  tibble(Model = "Linear Regression", RMSE = rmse_lm))
kable(rmse_results, caption = "Model Performance Tracking")
```

Linear regression improves substantially over the baseline by 
incorporating all eight features. However its core assumption, that features contribute independently and additively, is 
unlikely to hold perfectly in retail sales data. Promotions 
interact with day of week, store type interacts with 
assortment, and seasonal effects vary by store. These 
interaction effects motivate the use of tree-based models next.

### 5.4 - Decision Tree

A decision tree was applied as the first non-linear model. 
Unlike linear regression, decision trees partition the data 
into segments based on feature thresholds, capturing 
interaction effects that linear models cannot express. For 
example, the effect of a promotion may differ by store type 
or day of week, and a tree can learn these combinations 
naturally.

The tree depth was limited to 10 levels to prevent 
overfitting while still capturing meaningful patterns.

```{r model-tree, fig.height=6}
tree_model <- rpart(Sales ~ .,
                    data    = train_set,
                    method  = "anova",
                    control = rpart.control(maxdepth = 10))

# Plot only the top 3 levels for readability
# (full tree has 1,115 store labels that overwhelm the plot)
tree_simple <- rpart(Sales ~ .,
                     data    = train_set,
                     method  = "anova",
                     control = rpart.control(maxdepth = 3))

rpart.plot(tree_simple,
           main  = "Decision Tree Structure (Top 3 Levels)",
           type  = 4,
           extra = 101,
           tweak = 1.1)

tree_pred <- predict(tree_model, newdata = test_set)
tree_pred <- pmax(tree_pred, 0)

rmse_tree <- RMSE(test_set$Sales, tree_pred)
cat("Decision Tree RMSE:", rmse_tree, "\n")

rmse_results <- bind_rows(rmse_results,
  tibble(Model = "Decision Tree", RMSE = rmse_tree))
kable(rmse_results, caption = "Model Performance Tracking")
```

The decision tree is an improvement over baseline model, but Linear regression is so far the best one.

### 5.5 - Random Forest

Random Forest builds a large number of decision trees, each 
trained on a random subset of the data and features. The 
final prediction is the average across all trees. This 
ensemble approach reduces the variance of individual trees 
and typically produces the best predictive performance.

Due to computational constraints, the randomForest package cannot handle factor variables with more than 53 levels. Store (1,115 levels) was therefore converted to a numeric identifier. Additionally, a random sample of 50,000 observations was drawn from the training set to keep runtime manageable. The model was configured with 100 trees and a minimum node size of 25.

```{r model-rf}
set.seed(42)
train_rf <- train_set %>%
  mutate(Store = as.numeric(as.character(Store))) %>%
  sample_n(50000)

test_rf <- test_set %>%
  mutate(Store = as.numeric(as.character(Store)))

rf_model <- randomForest(
  Sales ~ .,
  data       = train_rf,
  ntree      = 100,
  mtry       = 3,
  nodesize   = 25,
  importance = TRUE
)

rf_pred <- predict(rf_model, newdata = test_rf)
rf_pred <- pmax(rf_pred, 0)                       
rmse_rf <- RMSE(test_set$Sales, rf_pred)    

rmse_results <- bind_rows(rmse_results,
  tibble(Model = "Random Forest", RMSE = rmse_rf))
kable(rmse_results, caption = "Model Performance Tracking")
```

The Random Forest did not outperform the baseline in this configuration. The sampling and tree constraints necessary for computational feasibility prevented the ensemble from learning store-level patterns as effectively as linear regression's fixed effects. Nevertheless, the model provides valuable diagnostic information through its feature importance analysis below.

### 5.6 - Feature Importance

One of the key advantages of Random Forest is its ability to 
quantify how much each feature contributes to predictive 
accuracy. The chart below shows the percentage increase in 
prediction error if each feature were randomly shuffled,
higher values indicate more important features.

```{r plot-importance}
importance_df <- data.frame(
  Feature    = rownames(importance(rf_model)),
  Importance = importance(rf_model)[, "%IncMSE"]
) %>%
  arrange(desc(Importance))

ggplot(importance_df,
       aes(x = reorder(Feature, Importance),
           y = Importance)) +
  geom_col(fill = "steelblue", color = "black") +
  coord_flip() +
  ggtitle("Random Forest - Feature Importance") +
  xlab("Feature") +
  ylab("% Increase in MSE if Removed")
```

The feature importance chart confirms the findings from the 
EDA, Store identity and promotional activity are the 
dominant drivers of daily sales, followed by temporal 
features capturing day of week and seasonal patterns.
Please note that here Store appears mid-ranked because it was converted to a numeric 
identifier for the Random Forest, as a continuous number, the 
model cannot learn individual store profiles the way linear 
regression's factor encoding does, which partly explains why 
the Random Forest underperformed.

## 6 - Results

```{r final-results}
kable(rmse_results, caption = "Final Model Performance Summary")
```

```{r results-chart}
rmse_results %>%
  ggplot(aes(x = reorder(Model, RMSE), y = RMSE, fill = RMSE)) +
  geom_col(color = "black") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_gradient(high = "tomato", low = "steelblue") +
  ggtitle("Model Comparison: RMSE (Lower is Better)") +
  xlab("") + ylab("RMSE (Euros)") +
  theme(legend.position = "none")
```

Linear Regression emerged as the best-performing model with an 
RMSE of approximately 1,370 euros, representing a 28% reduction 
in prediction error compared to the baseline. The Decision Tree 
also outperformed the baseline, achieving an RMSE of 
approximately 1,599. The Random Forest, despite being the most 
sophisticated algorithm tested, did not surpass the baseline due 
to computational constraints that limited its ability to learn 
store-level patterns.

The dominance of Linear Regression in this project carries an 
important practical insight. With 1,115 unique stores, the 
single most valuable piece of information for predicting daily 
sales is *which store* is being forecast. The baseline model, 
which uses only store identity, already explains enough variance 
to reach an RMSE of 1,916, and linear regression's main 
advantage over tree-based alternatives is its ability to encode 
each store as a separate factor level, essentially learning a 
unique baseline for each store. The remaining features 
(promotions, seasonality, store type) then adjust this baseline 
up or down. This simple, interpretable structure proved more 
effective than the constrained tree-based alternatives, where 
Store had to be treated as a numeric variable and lost its 
categorical power.

From a deployment perspective, the linear model also offers 
practical advantages: it is fast to train, produces 
interpretable coefficients, and is straightforward to update 
as new data arrives, all desirable properties for an 
operational demand forecasting system.

\newpage

## 7 - Conclusion

This project successfully built a demand forecasting model for 
Rossmann drug stores using historical sales data across 1,115 
stores over two and a half years. Through systematic exploratory 
analysis, we identified promotions, day of week, seasonality, 
store type, and competitor proximity as the key drivers of daily 
sales. The final linear regression model achieved an RMSE of 
approximately 1,370 euros, reducing prediction error by 28% 
compared to a naive store-average baseline.

**Key Findings**

The most impactful discovery was the overwhelming importance of 
store identity as a predictor. Each store has a fundamentally 
different sales profile shaped by its location, format, and 
customer base. Promotions were the second strongest signal, 
consistently lifting sales across all store types. The 
competitor distance analysis revealed a nuanced dynamic: stores 
near competitors attract more customers but generate smaller 
basket sizes, a finding with direct implications for inventory 
replenishment strategy.

**Limitations**

The project has three notable limitations. First, the Random 
Forest model was constrained by computational resources, the 
`randomForest` package cannot natively handle categorical 
variables with more than 53 levels, and the full dataset 
required sampling to remain tractable. A more powerful 
implementation (such as the `ranger` package or gradient 
boosting via `xgboost`) could potentially outperform linear 
regression. Second, the model does not capture store-level 
interactions with other features, for example, promotions 
may have different effects at different store types, which an 
interaction term or more flexible model could address. Third, 
the model has no mechanism for incorporating external data 
such as weather, local events, or macroeconomic conditions, 
all of which influence retail demand.

**Future Work**

Future iterations could explore gradient boosting (XGBoost), 
which handles large categorical features efficiently and 
typically achieves state-of-the-art performance in tabular 
prediction tasks. Additionally, incorporating interaction 
terms between Store and Promo, or between StoreType and 
DayOfWeek, could capture the non-linear dynamics that the 
current additive model misses. Finally, transitioning from 
point predictions to probabilistic forecasts would provide 
confidence intervals, essential for real-world inventory 
planning where the cost of overstocking differs from the cost 
of stockouts.

## References

- Rossmann Store Sales dataset. Kaggle competition hosted by 
  Rossmann. Available at: 
  https://www.kaggle.com/c/rossmann-store-sales/data
